---
title: "The auxiliary commands which can help to the users"
author: "Selcen Ari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{auxiliary_commands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning= FALSE, message=FALSE}
library(ceRNAnetsim)
```

In the other package vignettes, usage of ceRNAnetsim is explained in details. But in this vignette, some of commands which facitate to use of other vignette.

```{r}
data("TCGA_E9_A1N5_tumor")
data("TCGA_E9_A1N5_normal")
data("mirtarbasegene")
```

## Selection of HIST1H3H gene at vignette [How does the system behave in mirtarbase dataset without interaction factors?](https://selcenari.github.io/ceRNAnetsim/articles/mirtarbase_example.html)

```{r}
TCGA_E9_A1N5_mirnanormal %>%
  inner_join(mirtarbasegene, by= "miRNA") %>%
  inner_join(TCGA_E9_A1N5_normal, 
             by = c("Target"= "external_gene_name")) %>%
  select(Target, miRNA, total_read, gene_expression) %>%
  distinct() -> TCGA_E9_A1N5_mirnagene
```


```{r}
TCGA_E9_A1N5_tumor%>%
  inner_join(TCGA_E9_A1N5_normal, by= "external_gene_name")%>%
  select(patient = patient.x, 
         external_gene_name, 
         tumor_exp = gene_expression.x, 
         normal_exp = gene_expression.y)%>%
  distinct()%>%
  inner_join(TCGA_E9_A1N5_mirnagene, by = c("external_gene_name"= "Target"))%>%
  filter(tumor_exp != 0, normal_exp != 0)%>%
  mutate(FC= tumor_exp/normal_exp)%>%
  filter(external_gene_name== "HIST1H3H")

#HIST1H3H: interacts with various miRNA in dataset, so we can say that HIST1H3H is non-isolated competing element and increases to 30-fold.

```

## Selection of ACTB gene at vignette [How does the system behave in mirtarbase dataset without interaction factors?](https://selcenari.github.io/ceRNAnetsim/articles/mirtarbase_example.html)

```{r}
TCGA_E9_A1N5_tumor%>%
  inner_join(TCGA_E9_A1N5_normal, by= "external_gene_name") %>%
  select(patient = patient.x, 
         external_gene_name, 
         tumor_exp = gene_expression.x, 
         normal_exp = gene_expression.y) %>%
  distinct() %>%
  inner_join(TCGA_E9_A1N5_mirnagene, 
             by = c("external_gene_name"= "Target")) %>%
  filter(tumor_exp != 0, normal_exp != 0) %>%
  mutate(FC= tumor_exp/normal_exp) %>%
  filter(external_gene_name == "ACTB")

#ACTB: interacts with various miRNA in dataset, so ACTB is not isolated node in network and increases to 1.87-fold.

```

