---
title: "Basic Example for Usage of ceRNAnetsim"
author: "Selcen Ari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message= FALSE, warning=FALSE}
library(ceRNAnetsim)
```

You can obtain your own dataset with basic commands as following:

```{r, message= FALSE, warning=FALSE}
# datapasta::tribble_paste()
gene_expression <- tibble::tribble(
                  ~competing, ~Competing_expression,
                     "Gene1",                 10000,
                     "Gene2",                 10000,
                     "Gene3",                  5000,
                     "Gene4",                 10000,
                     "Gene5",                  5000,
                     "Gene6",                 10000
                  )
```

```{r}
gene_expression
```

```{r, message= FALSE, warning=FALSE}
# datapasta::tribble_paste()
mirna_expression <- tibble::tribble(
                             ~miRNA, ~miRNA_expression,
                             "Mir1",              1000,
                             "Mir2",              2000
                             )
```

```{r}
mirna_expression
```

Third table should contain miRNA:gene interactions. This table can be as simple as

```{r, message= FALSE, warning=FALSE}
#datapasta::tribble_paste()
interaction_simple <- tibble::tribble(
                     ~competing, ~miRNA,
                        "Gene1", "Mir1",
                        "Gene2", "Mir1",
                        "Gene3", "Mir1",
                        "Gene4", "Mir1",
                        "Gene4", "Mir2",
                        "Gene5", "Mir2",
                        "Gene6", "Mir2"
                     )
```


```{r}
interaction_simple
```

```{r}
interaction_simple%>%
  inner_join(gene_expression, by = "competing")%>%
  inner_join(mirna_expression, "miRNA")->basic_data
```

## What does ceRNAnetsim do?

ceRNAnetsim processes your dataset as graph object and simulates competing behaviours of targets when steady-state is perturbed via one or more change. Let's run three steps

## 1. Handle basic dataset

```{r}
#Convertion of dataset to graph.

priming_graph(basic_data, competing_count = Competing_expression, miRNA_count =miRNA_expression)
```

## 2. Trigger a change

```{r}

priming_graph(basic_data, competing_count = Competing_expression, miRNA_count =miRNA_expression)%>%
  update_how(node_name = "Gene2", how=2)

```

You can see the current count of Gene2 node is 20000 and its change is specified as "Up" in changes_variable column of node data of graph.

## 3. Simulate the changes in graph

```{r}

priming_graph(basic_data, competing_count = Competing_expression, miRNA_count =miRNA_expression)%>%
  update_how(node_name = "Gene2", how=2)%>%
  simulate(cycle = 5)

```

## In addition

ceRNAnetsim also provides the simulation of gene knockdown in the network. In normal conditions, when a gene is up or down regulated, it is considered that amounts of gene transcripts change depended on interactions. But, the transcripts of the gene are not observed in the system when it is knocked down.

In this context, we have developed 'gene_knockdown' and 'simulate_knockdown' functions. You can utilize these functions as following:

```{r}
data("midsamp")

midsamp %>%
   priming_graph(competing_count = Gene_expression, 
                 miRNA_count = miRNA_expression, aff_factor = c(seeds, Energy), deg_factor = targeting_region)%>%
  gene_knockdown("Gene4")%>%
  simulate_knockdown(cycle = 10, "Gene4")


```

Briefly, ceRNAnetsim utilizes the change(s) as trigger and provides regulation of targets according to miRNA:target and target:total target ratios. 

See the other vignettes for more information.

