---
title: "A TCGA dataset application"
author: "Selcen Ari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An TCGA dataset application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, echo=FALSE, message= FALSE, warning=FALSE}
library(svglite)
```

# Introduction

This vignette is about the integration of gene and miRNA pairs and their expression dataset and analysis. The sample dataset in this demonstration, which contains human miRNA:target pairs, was retrieved from [miRTarBase](http://mirtarbase.mbc.nctu.edu.tw/php/download.php) website (Release 7.0).

## miRNA:target pairs

```{r, warning= FALSE, message=FALSE}
# install.packages("devtools")
# devtools::install_github("selcenari/ceRNAnetsim")

library(ceRNAnetsim)
data("mirtarbasegene")
head(mirtarbasegene)
```

> **NOTE** if the mirna:target dataset includes miRNA genes as targets, the `priming_graph()` function can fail. Because, the function define to miRNAs and targets without distinguishing between uppercase or lowercase. 


<!-- #TODO we should fix this issue, labeling "competing" should not be regex based  #TODO priming_graph is fixed, it's not regex based anymore but this issue, miRNA affecting miRNA gene might need care, we should advice user about naming mature miRNA and miRNA gene in a distinguishable way -->

## Gene expression in normal and tumor samples

The gene and mirna expression counts of patient barcoded with TCGA-E9-A1N5 is retrieved from TCGA via `TCGAbiolinks` package from `Bioconductor`. The instructions of retrieving data can be found at `TCGAbiolinks` manual. 

For this step you don't have to use TCGA data, any other source or package can be utilized.

```{r}
data("TCGA_E9_A1N5_normal")
head(TCGA_E9_A1N5_normal)
```

```{r}
data("TCGA_E9_A1N5_tumor")
head(TCGA_E9_A1N5_tumor)
```

## miRNA expression data

```{r}
data("TCGA_E9_A1N5_mirnatumor")
head(TCGA_E9_A1N5_mirnatumor)
```

```{r}
data("TCGA_E9_A1N5_mirnanormal")
head(TCGA_E9_A1N5_mirnanormal)
```

Here's the summary of size of each dataset

|Dataset name| Number of rows |
|------------|----------------|
| `mirtarbasegene`           | `r nrow(mirtarbasegene)`           |
| `TCGA_E9_A1N5_normal`      | `r nrow(TCGA_E9_A1N5_normal)`      |
| `TCGA_E9_A1N5_tumor`       | `r nrow(TCGA_E9_A1N5_tumor)`       |
| `TCGA_E9_A1N5_mirnanormal` | `r nrow(TCGA_E9_A1N5_mirnanormal)` |
| `TCGA_E9_A1N5_mirnatumor`  | `r nrow(TCGA_E9_A1N5_mirnatumor)`  |

# Integrating data

All of these datasets are integrated using the code below resulting in  miRNA:target dataset that contains miRNA and gene expression values.

```{r}
TCGA_E9_A1N5_mirnanormal %>%
  inner_join(mirtarbasegene, by= "miRNA") %>%
  inner_join(TCGA_E9_A1N5_normal, 
             by = c("Target"= "external_gene_name")) %>%
  select(Target, miRNA, total_read, gene_expression) %>%
  distinct() -> TCGA_E9_A1N5_mirnagene
```


```{r, echo=FALSE}

TCGA_E9_A1N5_mirnagene%>%
  group_by(Target, miRNA)%>%
  count()%>%
  filter(n==2)

TCGA_E9_A1N5_mirnagene %>%    
  group_by(Target) %>%        
  mutate(gene_expression= max(gene_expression)) %>%
  distinct() %>%
  ungroup() -> TCGA_E9_A1N5_mirnagene


```

<!-- #TODO in the code above what is the purpose of TCGA_E9_A1N5_mirnagene %>% group_by(Target) %>% mutate(gene_expression= max(gene_expression)) %>% distinct() %>% ungroup() , are there multiple expressions per gene/miRNA--> 


<!-- yes, I did not understand why it occur when I processed the dataset. In chunk 9, you can see that dataset contains 2 expression values for 3 genes. Furthermore, you can find the integration process of this dataset in article repository and see that I have made merging process through ensembl gene id. Do you have any advice about that? -->

```{r}
head(TCGA_E9_A1N5_mirnagene)
```

When we compared the two gene expression dataset of TCGA-E9A1N5 patient, and selected a gene which has 30-fold increased expression, (gene name: HIST1H3H), this gene node will be used in the example. Note that the selected node must not be isolated one. If the an isolated node is selected the change in expression will not propagate in network. (You can see commands for node selection in the vignette [The auxiliary commands which can help to the users](https://selcenari.github.io/ceRNAnetsim/articles/auxiliary_commands.html))

Optionally, you can filter the low expressed gene nodes because they are not effective elements.

```{r}
TCGA_E9_A1N5_mirnagene%>%
  filter(gene_expression > 10)->TCGA_E9_A1N5_mirnagene
```

The analysis is performed based on amounts of miRNAs and targets as seen. Firstly, we tried to find optimal iteration for the network when simulation start with *HIST1H3H* node.

```{r, fig.height=4, fig.width=5, fig.align='center', warning=FALSE}
TCGA_E9_A1N5_mirnagene %>% 
  priming_graph(competing_count = gene_expression, 
                miRNA_count = total_read) %>% 
  update_how(node_name = "HIST1H3H", how =30) %>% 
  simulate(20) %>% 
  find_iteration(plot=TRUE)
```

The graph was shown that the change in expression level of HIST1H3H results in weak perturbation efficiency, despite 30-fold change. The results of the simulation was shown in following:

```{r}
TCGA_E9_A1N5_mirnagene %>%
  priming_graph(competing_count = gene_expression, 
                miRNA_count = total_read) %>%
  update_how("HIST1H3H", 30) %>%
  simulate(10)%>%
  as_tibble()%>%
  mutate(FC= count_current/initial_count)%>%
  arrange(desc(FC))
```

And then, we tried to simulate the network with the gene which has higher expression amount. For this, we selected *ACTB* node as shown in [The auxiliary commands which can help to the users](vignettes/auxiliary_commands.Rmd)


```{r, fig.height=5, fig.width=6, fig.align='center', warning=FALSE}
TCGA_E9_A1N5_mirnagene %>% 
  priming_graph(competing_count = gene_expression, 
                miRNA_count = total_read) %>% 
  update_how(node_name = "ACTB", how =1.87) %>% 
  simulate(10) %>% 
  find_iteration(plot=TRUE)
```


```{r}
TCGA_E9_A1N5_mirnagene %>%
  priming_graph(competing_count = gene_expression, 
                miRNA_count = total_read) %>%
  update_how("ACTB", 1.87) %>%
  simulate(10)%>%
  as_tibble()%>%
  mutate(FC= count_current/initial_count)%>%
  arrange(desc(FC))
```

# The sum of two conditions:

In a real biological sample, we tested perturbation efficiencies of two genes;
* one with low expression but high fold change (HIST1H3H, 30-fold increase in tumor) 
* another one with high expression but small change in expression level (ACTB, 1.87-fold increase in tumor)

With these two samples, we have been obtained that expression values of genes, rest of the perturbed gene, changed slightly

Despite high fold change, former gene caused little perturbation. On the contrary, high expressing gene with very low fold increase in tumor caused greater perturbation in the network.

Thus, if the perturbed node has lower target:total target ratio in group or groups, the efficiency of it can be weak, or vice versa. The efficiency of ACTB gene may be high for this reason, in comparison with HIST1H3H perturbation. In fact, it has been observed that ACTB has not strong perturbation efficiency too. This could be arisen from low miRNA:target ratio or ineffective target nodes which have very low expression levels.
