---
title: "An TCGA dataset application"
author: "Selcen Ari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An TCGA dataset application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, echo=FALSE, message= FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(stringi)
library(tidygraph)
library(igraph)
library(ggraph)
library(tidyverse)
library(ggplot2)
library(svglite)
library(devtools)

```

# Introduction

This vignette is about the integration of gene miRNA pairs and their expression dataset and analysis. Let's try to analysis of these dataset with `ceRNAnetsim`. First of all, obtain the datasets for analysis:

- The homo sapiens miRNA:target dataset is obtained via directly downloading from miRTarBase website. The dataset that is found in the working directory is processed as following:

```{r, warning= FALSE, warning=FALSE}
#install ceRNAnetsim

#install.packages("devtools")
#devtools::install_github("selcenari/ceRNAnetsim")


library(ceRNAnetsim)
```
NOTE that if the mirna:target dataset includes miRNA genes as targets, the priming_graph() function can get fail. Because, the function define to miRNAs and targets without distinguishing between uppercase or lowercase.

```{r}
data("mirtarbasegene")

head(mirtarbasegene)
```

- The gene and mirna expression counts of patient barcoded with TCGA-E9-A1N5 is obtained via `TCGAbiolinks` package from `Bioconductor`. The instructions of `TCGAbiolinks` can be followed. (For this step an other method (package) can be used.)

```{r}
data("TCGA_E9_A1N5_normal")
head(TCGA_E9_A1N5_normal)


data("TCGA_E9_A1N5_tumor")
head(TCGA_E9_A1N5_tumor)

```

```{r}
data("TCGA_E9_A1N5_mirnatumor")

head(TCGA_E9_A1N5_mirnatumor)

data("TCGA_E9_A1N5_mirnanormal")

head(TCGA_E9_A1N5_mirnanormal)

```

- All of these datasets are integrated. So, we obtains the miRNA:target dataset that contains miRNA and gene expression values.

```{r}

TCGA_E9_A1N5_mirnanormal%>%
  inner_join(mirtarbasegene, by= "miRNA")%>%
  inner_join(TCGA_E9_A1N5_normal, by = c("Target"= "external_gene_name"))%>%
  select(Target, miRNA, total_read, gene_expression)%>%
  distinct()->TCGA_E9_A1N5_mirnagene


TCGA_E9_A1N5_mirnagene%>%
  group_by(Target)%>%
  mutate(gene_expression= max(gene_expression))%>%
  distinct()%>%
  ungroup()->TCGA_E9_A1N5_mirnagene
```

- We compare the two gene expression dataset of TCGA-E9A1N5 patient. So, we selected the gene which increased to 30 fold. (This gene node have been selected as an example.) Note that the selected node must be non-isolated one. If the isolated node selects, the system can not be affected from this distrubtion.

```{r}

TCGA_E9_A1N5_tumor%>%
  inner_join(TCGA_E9_A1N5_normal, by= "external_gene_name")%>%
  select(patient = patient.x, external_gene_name, tumor_exp = gene_expression.x, normal_exp = gene_expression.y)%>%
  distinct()%>%
  inner_join(TCGA_E9_A1N5_mirnagene, by = c("external_gene_name"= "Target"))%>%
  filter(tumor_exp != 0, normal_exp != 0)%>%
  mutate(FC= tumor_exp/normal_exp)%>%
  arrange(desc(FC))

#HIST1H3H : non-isolated gene. 30 FC.
```


The analysis is performed based on amounts of miRNAs and targets as seen. Firstly, we tried to find optimal iteration for the network when simulation start with *HIST1H3H* node.

```{r, fig.height=4, fig.width=5, fig.align='center', warning=FALSE}
iteration_graph(as.data.frame(TCGA_E9_A1N5_mirnagene), competing_count = gene_expression, miRNA_count = total_read, node_name = "HIST1H3H", .iter = 10, how = 30)
```


The graph was shown that the change of HIST1H3H results in weak perturbation efficiency, although strong change. The results of the simulation was shown in following:

```{r}
as.data.frame(TCGA_E9_A1N5_mirnagene)%>%
  priming_graph(competing_count = gene_expression, miRNA_count = total_read)%>%
  update_nodes(once = TRUE)%>%
  update_how("HIST1H3H", 30)%>%
  update_nodes()%>%
  simulate(10)

```


- And then, we tried to simulate the network with the gene which has higher expression amount. For this, we selected *ACTB* node as shown in following:

```{r}

TCGA_E9_A1N5_tumor%>%
  inner_join(TCGA_E9_A1N5_normal, by= "external_gene_name")%>%
  select(patient = patient.x, external_gene_name, tumor_exp = gene_expression.x, normal_exp = gene_expression.y)%>%
  distinct()%>%
  inner_join(TCGA_E9_A1N5_mirnagene, by = c("external_gene_name"= "Target"))%>%
  filter(tumor_exp != 0, normal_exp != 0)%>%
  mutate(FC= tumor_exp/normal_exp)%>%
  arrange(desc(normal_exp))

# ACTB 1.87 fold
```


```{r, fig.height=5, fig.width=6, fig.align='center', warning=FALSE}
iteration_graph(as.data.frame(TCGA_E9_A1N5_mirnagene), competing_count = gene_expression, miRNA_count = total_read, node_name = "ACTB", .iter = 10, how = 1.87)
```


```{r}
as.data.frame(TCGA_E9_A1N5_mirnagene)%>%
  priming_graph(competing_count = gene_expression, miRNA_count = total_read)%>%
  update_nodes(once = TRUE)%>%
  update_how("ACTB", 1.87)%>%
  update_nodes()%>%
  simulate(10)
```


# The comparison of two conditions:

- When the gene that has low expression level with high changing was used for perturbation of network, it displayed low efficiency with comparison to the gene which has low changing in higher expression level.

- If the perturbed node has lower target:total target ratio in group or groups, the efficiency of it can be weak, or vice versa. The efficiency of ACTB gene may be high for this reason. But, ACTB has not strong perturbation efficiency too. This could be arisen from low miRNA:target ratio or impotent target nodes which have very low expression levels.

